---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

### Load Data
```{r}
unzip("activity.zip")
activity_raw <- read.csv("activity.csv")
```

### Process Data

The data was processed by removing incomplete cases, then coercing the date 
column into Date objects. 

```{r}
activity_proc <- activity_raw[complete.cases(activity_raw),]
activity_proc$date <- as.Date(activity_proc$date)
```

The daily step total was calculated by summing the steps measured on the same
date.
```{r}
daily_total_steps <- tapply(activity_proc$steps,activity_proc$date, sum)
```



## What is mean total number of steps taken per day?
### Histogram of Total Number of Steps Taken Each Day

The histogram shows the daily step total frequencies. The observations
are grouped into 32 bins, which provides a balance of granularity and summary.
```{r}
library(ggplot2)
max_steps_in_day <- max(daily_total_steps)
qplot(daily_total_steps, 
      geom = "histogram", 
      binwidth = max_steps_in_day/32,
      xlab = "Daily Step Total",
      ylab = "Count",
      main = "Frequency of Daily Step Totals")
```

### Report Mean and Median of Total Number of Steps Taken per Day

The mean total number of steps per day is 
```{r}
mean_daily_total_steps <- mean(daily_total_steps)
mean_daily_total_steps
```

The median total number of steps per day is 
```{r}
median_daily_total_steps <- median(daily_total_steps)
median_daily_total_steps
```

## What is the average daily activity pattern?

### Time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

#### Process Data

This section will reuse the processed data created in the previous section.

The average number of steps in each interval is calculated by summing the number
of steps in each interval and dividing by the number of days. If a day does not
contain any valid values, then it is not be counted in the number of days used 
for the average. The first day in the data set, 2012-10-01, in not included for 
this reason.

```{r}
total_steps_in_interval <- tapply(activity_proc$steps,activity_proc$interval, sum)
num_of_days <- length(as.factor(activity_proc$date))

avg_steps_in_interval <- total_steps_in_interval/num_of_days

## Convert to a factor to extract all the unique values
interval_levels <- attr(as.factor(activity_proc$interval),"levels")
## Convert to numeric so the plot is sorted correctly
interval_levels <- as.numeric(interval_levels)

interval_avg_steps <- data.frame(interval_levels, avg_steps_in_interval)

qplot(interval_levels, 
      avg_steps_in_interval, 
      interval_avg_steps,
      geom = "line",
      xlab = "Interval", 
      ylab = "Average Number of Steps",
      main = "Average Number of Steps in Each Interval")

```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
index_max <- which.max(x = interval_avg_steps$avg_steps_in_interval)
interval_avg_steps$interval_levels[index_max]
```

## Imputing missing values
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
num_missing_values <- nrow(activity_raw[is.na(activity_raw),])
num_missing_values
```



## Are there differences in activity patterns between weekdays and weekends?
